Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-06 Manuscript/manuscript.tex
Update text, I am now done with the affinity section for now

According to ECC, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a dielectric continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of force field for ions \cite{kohagen14,kohagen16,??}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and 
improves the model performance against experimental scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e, lipids in our case. 
Unlike the total charge of molecules, atomic partial charges within each molecule are not experimental observables.
Although there have been proposed several schemes for the assignment of partial charges of molecules, 
methods targetting at fitting of electrostatic potentials are 
most popular among classical fixed charge force fields for biomolecules. \cite{Hu2007}
Currently the most commonly employed scheme is the RESP scheme~\cite{RESP_paper,Singh1984}, 
which fits the electrostatic potential from quantum mechanical calculations. 
The resulting set of fixed partial atomic charges is hence 
the best fit to electrostatic potentials of 
an ensemble of conformations of a molecule. 
Such a physical meaning is especially well suited for the application of ECC, 
which targets at improving electrostatic interactions. 
Depending to the details, partial charge calculations may also partially
include some of the solvent electronic polarizability effects. 
Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges
does not have to be striclty $f_q = \epsilon _{el} ^{-1/2}$, 
but rather lies between this value (no electronic polarizability in the charge calculation)
and $1$ (full electronic polarizability already included in the charge calculation).
%
% This is not really important for the paper:
%
%However, the RESP method has several remedies;  %and these are specific to RESP
%RESP charges show a dependence on the orientation and position
%and it does not provide numerically stable partial charges of burried atoms;~\cite{Hu2007}
%the RESP-derived values depend, for example, on the conformational
%sampling of the molecules and description of solvent 
%in the calculations~\cite{RESP_paper, Singh1984, Hu2007, maciejewski14}. 
%
%Hirshfeld-E partitioning: \cite{Verstraelen2013}
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~~\cite{RESP_paper,Singh1984}),
%
%

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

@@

The headgroup P-N vector angle with respect to the membrane normal
is also shown in Fig.~\ref{OrderParameterCHANGESsurf}
as a function of cationic surfactant mole fraction.
The headgroup orients more parallel to the membrane
normal with increasing amount of bound cations, as expected~\cite{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the ECC-lipid model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
Therefore we use the data in Fig.~\ref{OrderParameterCHANGESsurf}
to suggest a linear relation between $\alpha$-carbon order parameter
and average P-N vector angle changes $\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}^{\alpha}$.
This can be used to estimate the change in P-N vector angle
from experiments measuring changes in $\alpha$-carbon order parameter. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}
For example, $\Delta S_{\rm CH}^{\alpha}=0.018$ measured due to the
addition of 66.8 $\mu$M concentration of Mellittin in bulk water \cite{kuchinka89}
gives $\Delta \Theta_{\rm P-N} = 2.5^{\circ}$.
\todo{I looked this roughly from the figure, should be calculated from 
the actual relation if we decide to put this.}

@@

Significant improvement is achieved when the developed ECC-lipid model 
is used together with ECC-ions. The headgroup order parameter changes
as a function of CaCl$_2$ concentration show a good agreement between
ECC-lipid model and experiments in Fig. \ref{OrderParameterCHANGESnewMODELS}.
As discussed in previous section, also the response of headgroup order
parameters to the bound charge is in good agreement with the experiments in this
model. Thus, we conclude that the ECC-lipid model with ECC-ion parameters
gives a realistic binding affinity for the Na$^+$ and Ca$^{2+}$ binding
in POPC bilayer.

Based on the ion density profiles along membrane normal
in Fig.~\ref{fig:cacl-dens}, the Ca$^{2+}$ binding seems to
be stronger in Lipid14 model with Dang and ECC-ions.
This is in agreement 
This is quantified by calculating the relative surface
excess from Eq.~\ref{surfexcess}. This gives 
$\Gamma_i^w = 0.07 nm^{-3}$ for ECC-lipid model simulated with ECC-ions,
which is significantly smaller than $\Gamma_i^w = XX nm^{-3}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 nm^{-3}$ with Dang ions.
\todo{Surface excess for Lipid14 with Aqvist yet to be added.} \\
\todo{Surface excess units are not correct. Maybe the numbers needs to be checked also?} \\
Rough estimates for the free energy difference between bound and unbound cation
can made from
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

Since ECC--lipid model gives a good agreement with experiments for the response
of headgroup order parameters to the bound charge as well as to the CaCl$_2$
concentration,

The good agreement with experiments indicates
that the ECC corrected lipid model has a sufficient accuracy for realistic
studies of details of ions binding to lipid bilayers.

>>>>

According to ECC, electronic
polarizability can be implicitly included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a dielectric continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of force field for ions \cite{kohagen14,kohagen16,??}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and 
improves the model performance against experimental scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e, lipids in our case. 
Unlike the total charge of molecules, atomic partial charges within each molecule are not experimental observables.
\todo{Text should be clarified from here to the end of the paragraph}
Although there have been proposed several schemes for the assignment of partial charges of molecules, 
methods targetting at fitting of electrostatic potentials are 
most popular among classical fixed charge force fields for biomolecules. \cite{Hu2007}
Currently the most commonly employed scheme is the RESP scheme~\cite{RESP_paper,Singh1984}, 
which fits the electrostatic potential from quantum mechanical calculations. 
The resulting set of fixed partial atomic charges is hence 
the best fit to electrostatic potentials of 
an ensemble of conformations of a molecule. 
Such a physical meaning is especially well suited for the application of ECC, 
which targets at improving electrostatic interactions. 
Depending to the details, partial charge calculations may also partially
include some of the solvent electronic polarizability effects. 
Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges
does not have to be striclty $f_q = \epsilon _{el} ^{-1/2}$, 
but rather lies between this value (no electronic polarizability in the charge calculation)
and $1$ (full electronic polarizability already included in the charge calculation).
%
% This is not really important for the paper:
%
%However, the RESP method has several remedies;  %and these are specific to RESP
%RESP charges show a dependence on the orientation and position
%and it does not provide numerically stable partial charges of burried atoms;~\cite{Hu2007}
%the RESP-derived values depend, for example, on the conformational
%sampling of the molecules and description of solvent 
%in the calculations~\cite{RESP_paper, Singh1984, Hu2007, maciejewski14}. 
%
%Hirshfeld-E partitioning: \cite{Verstraelen2013}
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~~\cite{RESP_paper,Singh1984}),
%
%

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

@@

The headgroup P-N vector angle with respect to the membrane normal
is also shown as a function of cationic surfactant mole fraction
in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more parallel to the membrane
normal with increasing amount of bound cations, as expected~\cite{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the ECC-lipid model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
Therefore we use the data in Fig.~\ref{OrderParameterCHANGESsurf}
to suggest a linear relation between $\alpha$-carbon order parameter
and average P-N vector angle changes $\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}^{\alpha}$.
This can be used to estimate the change in P-N vector angle
from experiments measuring changes in $\alpha$-carbon order parameter. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}
For example, $\Delta S_{\rm CH}^{\alpha}=0.018$ measured due to the
addition of 66.8 $\mu$M concentration of Mellittin in bulk water \cite{kuchinka89}
gives $\Delta \Theta_{\rm P-N} = 2.5^{\circ}$.
\todo{I looked this roughly from the figure, should be calculated from 
the actual relation if we decide to put this.}

@@

Significant improvement can be achieved by using the ECC
approach also for lipids. The headgroup order parameter changes
as a function of CaCl$_2$ concentration from ECC-lipid model with ECC-ions
show a good agreement with experiments in Fig. \ref{OrderParameterCHANGESnewMODELS}.
As discussed in previous section, the model gives also
a good agreement with experiments for
the headgroup response to bound charge.
Thus, the model can used for more detailed analysis of the
binding affinity.

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 nm^{-3}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = XX nm^{-3}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 nm^{-3}$ with Dang ions.
\todo{Surface excess for Lipid14 with Aqvist yet to be added.} \\
\todo{Surface excess units are not correct. Maybe the numbers needs to be checked also?} \\
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

Since the lipid headgroup order parameter responses to the amount of bound charge
and to the aqueous ion concentrations are both in good agreement with experiments
in the ECC-lipid model with ECC-ion parameters, we consider the Na$^+$ and Ca$^{2+}$
binding affinities to be realistic in this model. On the other hand, the Ca$^{2+}$
binding affinity is overestimated by Lipid14 model when simulated with all the
tested ion models. Similar conclusions were previously made based only on the headrogup
order parameter data with aqueous cations~\cite{catte16}. However,
the discrepancies with experiments in previous work could partly arise also from
the inaccurate sensitivity of the headgroup to bound charge. Here we quantify this
effect and conclude that the improvement due to ECC is partly, but not completely,
caused by more realistic headgroup sensitivity to bound charge.
This indicates that the issue should be carefully considered also when
the electrometer concept is used to compare ion binding between experiments
and simulations wiht other models.

%The good agreement with experiments indicates
%that the ECC corrected lipid model has a sufficient accuracy for realistic
%studies of details of ions binding to lipid bilayers.
